Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2013-03-24 HRUIonGPU/jrnl_article.tex
Moved first part of beamspace section to background in tuffc article

\author{J. P. \AA{}sen, \IEEEmembership{Student Member,~IEEE,} J. I. Buskenes, C.-I. C. Nilsen, \IEEEmembership{Member,~IEEE,} A. Austeng, \IEEEmembership{Member,~IEEE,} and S. Holm, \IEEEmembership{Senior Member,~IEEE,} %<-this % stops a space
%\thanks{Manuscript received ...; accepted ...}
\thanks{Jon Petter \AA{}sen and Sverre Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway. Jo Inge Buskenes, Carl-Inge Colombo Nilsen, Andreas Austeng and Sverre Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
}

@@

\begin{abstract}
%\boldmath
Capon beamforming is associated with a high computational complexity, which limits its use as a real-time method in many applications. In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases.

For a three dimensional beamspace, processing rates supporting real-time cardiac ultrasound imaging are shown to be possible, meaning that processing rates higher than the image acquisition rate are obtained for a wide range of parameters. Image quality is investigated in videos of both a simulated phantom and an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element- and beamspace.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Most of this work focuses on the improved lateral resolution and contrast obtained with the Capon beamformer. However, some interesting trade-offs have also been introduced. In \cite{Synnevag2009}, Synnev\aa{}g et al. explains how image resolution can be maintained at higher framerates, with smaller probes, or for deeper penetration. Despite these benefits, widespread adoption of the Capon beamformer has not been seen due to its high computational complexity \cite{So2011}. In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To reach this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs) (found in almost all PCs), combined with novel methods to reduce the computational complexity as the number of array elements increases.

Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. However, the Capon beamformer is meant to work on complex data. Under the constraint of real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image \cite[Fig 9.]{Synnevag2009}. Complex arithmetic also requires more operations per sample, but the sampling frequency can also be reduced. The number of samples needed per image is therefore minimized. All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is close to the system bandwidth.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer. Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information on the Capon beamformer and the GPU computing model. In Section \ref{sec:meth} we go in detail on the parallel implementation of the Capon beamformer, before the beamspace method, and its GPU implementation, is presented in Section \ref{sec:bs}. Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.  Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

\section{Background}

\subsection{Capon Beamforming}

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n]. \label{eq:z}
\end{align}
The weight vector, or window, $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR). Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

As shown in Fig. \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. It is worth noting that the micro-steering caused by the complex weights is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

@@

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equation, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
Where $f(\cdot)$ represents the process of constructing $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

@@

Finally the Capon weight vector is formed from (\ref{eq:b}) as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
\end{align}
and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. This is formally stated as
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
&= 1/N_L \, (\vec{w}[n] *_{m} \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $*_m$ is a non-truncated convolution across elements.
Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate a weighting per element (\ref{eq:z_mv2}).

@@

@@

In this section we give a brief discussion of GPU computing, laying the basis for further discussion on parallel implementation strategies. As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g. OpenCL. The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU. While the CPU has few cores and little logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to achieve maximum throughput.

The CUDA compute model is based on the execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

An SM has a limited amount of local (near-core or shared) memory and registers. Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), each thread has to consume as few registers and as little local memory as possible. Each thread should also perform enough instructions per transferred byte in order to hide memory latency. Dividing the problem into enough threads will help to hide instruction latency. Global GPU memory transactions, in addition to CPU-GPU transfers, are slow and must be minimized. All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration. In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.

@@

\section{Parallel Capon}\label{sec:meth}
If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with (\ref{eq:b}) and (\ref{eq:diag}), we see that each calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach. Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.

In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
Along the time dimension we also find that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of local memory available per thread, we focus on subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$. The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is thereby increased. When sliding across subarrays, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is now increased to the maximum number of threads per block (currently 1024). However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.

@@

Transforming subarrays into beamspace yields subarrays that no longer overlap. This means that the spatial overlap previously exploited when calculating the sample covariance matrix in element space is no longer available. On the other hand, since $N_b$ should be small ($N_b < 32$), one thread per element in $\mat{\hat{R}}_\text{BS}$ is now an option. A kernel similar to the one described in \cite{Chen2011} has therefore been implemented. In addition, for small $N_b$ values, several covariance matrices are calculated per thread block to achieve high occupancy of the GPU.

@@

\newpage

>>>>

\author{Jon Petter \AA{}sen, \IEEEmembership{Student Member,~IEEE,} Jo Inge Buskenes, Carl-Inge Colombo Nilsen, \IEEEmembership{Member,~IEEE,} Andreas Austeng, \IEEEmembership{Member,~IEEE,} and Sverre Holm, \IEEEmembership{Senior Member,~IEEE,} %<-this % stops a space
\thanks{Manuscript received ...; accepted ...}
\thanks{J. P. \AA{}sen and S. Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Science and Technology, Trondheim, Norway.}
\thanks{J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
}

@@

\begin{abstract}
%\boldmath
Capon beamforming is associated with a high computational complexity, which limits its use as a real-time method in many applications. In this paper we present an implementation of the Capon beamformer that exhibits real-time performance when applied in a typical cardiac ultrasound imaging setting. To achieve this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs), combined with beamspace processing to reduce the computational complexity as the number of array elements increases.

For a three dimensional beamspace we show that processing rates supporting real-time cardiac ultrasound imaging are possible, meaning that images can be processed faster than the image acquisition rate for a wide range of parameters. Image quality is investigated in videos of both a simulated phantom, and in an \textit{in-vivo} cardiac dataset. These results show that Capon beamforming is feasible for cardiac ultrasound imaging, providing images with improved lateral resolution and no local shift-variance effects, both in element-space and beamspace.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

@@

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} beamforming \cite{Capon1969} has recently been applied to medical ultrasound imaging with promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Most of this work focuses on the improved lateral resolution and contrast obtained with the Capon beamformer. However, some interesting trade-offs have also been introduced. In \cite{Synnevag2009}, Synnev\aa{}g \textit{et al.} explains how image resolution can be maintained at higher framerates, with smaller probes, or for deeper penetration. Despite these benefits, widespread adoption of the Capon beamformer has not been seen due to its high computational complexity \cite{So2011}. In this paper we present an implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To achieve this performance we make use of the parallel processing power found in modern Graphics Processing Units (GPUs) (found in almost all PCs), combined with novel methods to reduce the computational complexity as the number of array elements increases.

Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produces real apodization weights based on real data. However, the Capon beamformer is meant to process complex data. Under the constraint of using real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image \cite[Fig~9.]{Synnevag2009}. Complex arithmetic leads to more operations per sample, but also allows the sampling frequency to be reduced. The number of samples needed per image can therefore be minimized. All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is near the system bandwidth.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012, Buskenes, Buskenes2013} we have applied a GPU-based Capon beamformer for both active sonar imaging and for cardiac ultrasound imaging. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. Capon beamforming in medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates that not even a high-end GPU can provide. Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming for medical ultrasound imaging have therefore been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. 
%The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

In the next section we present background information on Capon beamforming in both element-space and beamspace and the GPU computing model. In Section \ref{sec:meth} we recap our parallel implementation of the Capon beamformer, before the beamspace method, and its GPU implementation, is presented in Section \ref{sec:bs}. Benchmark results and resulting images are presented in Section \ref{sec:bench} and \ref{sec:images}, together with a discussion of trading image quality for speed in Section \ref{sec:trade}.  Finally we discuss the results in Section \ref{sec:dis}, and draw conclusions in Section \ref{sec:con}.

\section{Background}\label{background}

\subsection{Capon Beamforming}\label{sec:es-capon}

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n]. \label{eq:z}
\end{align}
The weight vector, or window, $\vec{w}$ is usually real in DAS beamforming, and is applied to trade resolution for a better signal-to-noise ratio (SNR). Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

As shown in Fig. \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a complex weight vector based on the received data.   The Capon beamformer produces weights that minimize the beamformer output power while maintaining unity gain in the focus direction. The result is that interference impinging from other directions is suppressed \cite{Synnevag2007}. Micro-steering caused by the complex weights is minute compared to the initial steering and focusing step, therefore the beamformer also works well for broadband ultrasound signals.

@@

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equations, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of temporal samples included, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
Where $f(\cdot)$ represents the process of constructing $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

@@

Finally the Capon weight vector is formed from (\ref{eq:b}) as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{1}^H\vec{b}[n]} \in \mathbb{C}^L,
\end{align}
and the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights. This is formally stated as
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l[n] \label{eq:z_mv}\\
&= 1/N_L \, (\vec{w}[n] * \vec{1}_{N_L})^H\vec{x}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
where $*$ is a non-truncated convolution across elements.
Note how we can either choose to sum all subarrays (\ref{eq:z_mv}), or accumulate a weighting per element (\ref{eq:z_mv2}).

@@

\subsection{Beamspace Processing}\label{bs-capon}
Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M} \label{eq:butler} \\ 
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler or normalized DFT matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.

For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of the beams contains almost all the received energy. Nilsen and Hafizovic concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space. Hence, for equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.

The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
\begin{align}
\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
\end{align}
where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ represent the covariance estimation process as defined in (\ref{eq:diag}).

In the remaining sections, the methods described in this section and Section \ref{sec:es-capon} will be referred to as beamspace Capon (BS-Capon) and element-space Capon (ES-Capon) respectively. The name Capon will refer to both methods.

@@

In this section we give a brief discussion of GPU computing, laying the basis for the discussions on parallel implementation strategies. The Capon beamformers have been implemented using the CUDA GPU framework by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g. OpenCL. Independent of framework, the key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU. While the CPU has few cores and lack logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to thereby achieve maximum throughput.

In CUDA programming a kernel function is executed over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications or additions, or both, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

An SM has a limited amount of shared memory (or near-core cache) and registers. Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), each thread has to consume as few registers and as little shared memory as possible. Each thread should also perform enough instructions per transferred byte in order to hide memory latency. Dividing the problem into enough threads will help to hide instruction latency. Global GPU memory transactions, in addition to CPU-GPU transfers, are slow and must be minimized. All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration. In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.

@@

\section{Parallel ES-Capon}\label{sec:meth}
In this section we will recap our parallel implementation of the ES-Capon beamformer as presented in \cite{asen2012, Buskenes}. This implementation was our starting point when the BS-Capon beamformer was implemented.

If the beamformer in (\ref{eq:z_mv}) is analyzed in combination with the matrix equation in (\ref{eq:b}) and the covariance estimation process in (\ref{eq:diag}), we see that the calculation of a set of Capon weights is independent across the outputs $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be divided further into a set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations as in (\ref{eq:b}}).\\ %\vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}.
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexities of these three steps are $O(N_KN_LL^2)$, $O(L^3)$, and $O(L)$, respectively. As $M$ (and thereby $L$) increases, the first two steps quickly become unmanageable, and real-time performance can only be achieved by exploring methods with reduced complexity.

@@

Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of the other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach. Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.

In \cite{asen2012, Buskenes} we demonstrated how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
Along the time dimension we also find that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$. However, since $N_K$ typically is small, and the GPU has a limited amount of shared memory available per thread, we focus on subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$.

The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is therefore increased. Calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore assigned to each diagonal in the upper triangle of $\mat{\hat{R}}$. For small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is now increased to the maximum number of threads per block (currently 1024). However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full GPU occupancy.

@@

@@

%\newpage
